---
title: "bivalency_analysis"
author: "Rohan Shah"
date: "7/15/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Introduction

This R Markdown file is meant to streamline the analysis of the bivalency data and keep it organized in one place. Different types of analyses will be in different sections with explanations as to why those are important. This section will contain only data loading and functions that are necessary for downstream analyses.

### Data loading

This chunk loads in both the HMD data and the bulk RNA-seq data, as well as the tidyverse and sleuth libraries.

```{r data_loader}
library(tidyverse)
library(sleuth)

load("data_loaded.RData")
load("sleuth_loaded_bivalency_bulk-rna-seq.RData")

#Redo the HMDs for 0-400bp about the TSS

naive_hmd = data.frame(Gene = naive$k4$Gene)
naive_hmd$k4 = naive$k4 %>% select(X0:X400) %>% rowMeans
naive_hmd$k9 = naive$k9 %>% select(X0:X400) %>% rowMeans
naive_hmd$k27 = naive$k27 %>% select(X0:X400) %>% rowMeans
naive_hmd$cis = naive$cis %>% select(X0:X400) %>% rowMeans
naive_hmd$trans = naive$trans %>% select(X0:X400) %>% rowMeans
naive_hmd$Gene = as.character(naive_hmd$Gene)

primed_hmd = data.frame(Gene = primed$k4$Gene)
primed_hmd$k4 = primed$k4 %>% select(X0:X400) %>% rowMeans
primed_hmd$k9 = primed$k9 %>% select(X0:X400) %>% rowMeans
primed_hmd$k27 = primed$k27 %>% select(X0:X400) %>% rowMeans
primed_hmd$cis = primed$cis %>% select(X0:X400) %>% rowMeans
primed_hmd$trans = primed$trans %>% select(X0:X400) %>% rowMeans
primed_hmd$Gene = as.character(primed_hmd$Gene)

npc_hmd = data.frame(Gene = npc$k4$Gene)
npc_hmd$k4 = npc$k4 %>% select(X0:X400) %>% rowMeans
npc_hmd$k9 = npc$k9 %>% select(X0:X400) %>% rowMeans
npc_hmd$k27 = npc$k27 %>% select(X0:X400) %>% rowMeans
npc_hmd$cis = npc$cis %>% select(X0:X400) %>% rowMeans
npc_hmd$trans = npc$trans %>% select(X0:X400) %>% rowMeans
npc_hmd$Gene = as.character(npc_hmd$Gene)

tss_range = seq(-3, 3, by = 0.05)
```

The next chunk loads in functions to make metagenes.

```{r metagene_functions}
metagene = function(dataset, range, genes){
  dataset_names = names(dataset)
  summ_df = data.frame(dist = range)
  
  for (i in dataset_names){
    summ_df[,i] = as.numeric(dataset[[i]] %>% filter(Gene %in% genes) %>% select(-1) %>% summarize(across(.fns = mean)))
  }
  
  return(summ_df)
}

metagene_plot = function(summ_df, ylims = NA, title = NA){
  summ_plt = summ_df %>% gather("mod", "hmd", -dist)
  plot_summary = ggplot(data = summ_plt, aes(x = dist, y = hmd, color = mod)) + geom_line() + theme_classic()
  
  if (!is.na(ylims)){ plot_summary = plot_summary + scale_y_continuous(limits = ylims) }
  
  if (!is.na(title)){ plot_summary = plot_summary + ggtitle(title) }
  
  print(plot_summary)
}
```

## First-Pass HMD Metagene Analysis


